---
title: "NHS Gender Inequality: Data Visualisation Project"
author: "Annisa Kumaladewi"
output:
  html_document:
    self_contained: true
---


### Data Preparation
```{r}
library(tidyverse)

df <- read_csv("UK Gender Pay Gap Data.csv")

# Convert "2017-2018" -> 2018 (end year), etc.
df2 <- df %>%
  mutate(
    year_end = as.integer(str_sub(Year, -4))  # last 4 chars
  ) %>%
  arrange(year_end)
```

### Exporting Fonts: Roboto Condensed
```{r}
library(showtext)
library(sysfonts)

font_add(
  family = "Miller Text",
  regular = "fonts/fonnts.com-Miller_Text_Roman.otf",
  bold = "fonts/fonnts.com-Miller_Text_Bold.otf",
  italic = "fonts/fonnts.com-Miller_Text_Italic.otf",
  bolditalic = "fonts/fonnts.com-Miller_Text_Bold_Italic.otf"
)

font_add(
  family = "Miller Display",
  regular = "fonts/fonnts.com-Miller_Display_Roman.otf",
  bold = "fonts/fonnts.com-Miller_Display_Bold.otf",
  italic = "fonts/fonnts.com-Miller_Display_Italic.otf",
  bolditalic = "fonts/fonnts.com-Miller_Display_Light.otf"
)

# Note: "regular" = Light weight (intentional)
font_add(
  family = "Franklin Gothic",
  regular = "franklin-gothic/light.ttf",
  bold = "franklin-gothic/bold.ttf"
)

showtext_auto()

```

### Data Viz Libraries
```{r}
library(ggrepel)
```

## Viz 1: Mean/Median Gender Pay Gap @ NHS Barts Trust
### Data Manipulation
```{r}
gpg_long <- df2 %>%
  select(
    year_end,
    mean_gap = GenderPayGap_HourlyPay_Mean_Percent,
    median_gap = GenderPayGap_HourlyPay_Median_Percent
  ) %>%
  pivot_longer(cols = c(mean_gap, median_gap),
               names_to = "metric",
               values_to = "gap_pct") %>%
  mutate(metric = recode(metric,
                         mean_gap = "Mean (hourly)",
                         median_gap = "Median (hourly)"))

# Labels placed at last year for each line e.g. 2017-2018 becomes 2018
label_df <- gpg_long %>%
  group_by(metric) %>%
  filter(year_end == max(year_end)) %>%
  ungroup() %>%
  mutate(label = str_replace(metric, " \\(hourly\\)", ""))  # "Mean", "Median"
```

#Styling helpers
```{r}
col_map <- c("Mean (hourly)" = "#fcaa2d", "Median (hourly)" = "#0ab276")
y_position <-seq(10.3, 20.3, by = 2)
y_breaks <- seq(10, 20, by = 2)
y_vals <- y_breaks

```

  # Shaded region label (upper right)
  annotate("text",
           x = 2020.55, y = 20.7,
           label = "Post-pandemic period",
           hjust = 0, vjust = 1,
           size = 4.2,
           color = "gray25") +

### Plotting
```{r}
p <- ggplot(gpg_long, aes(x = year_end, y = gap_pct, group = metric, color = metric)) +

  # Post-pandemic shading
  annotate(
    "rect",
    xmin = 2020, xmax = 2021,
    ymin = -Inf, ymax = 20,
    alpha = 0.4, fill = "#e0eaf2"
  ) +

  # Line below Post-Pandemic
  annotate(
    "segment",
    x = 2020, xend = 2021,
    y = 20, yend = 20,
    linewidth = 0.3,
    color = "#7a9baa"
  ) +

  # Post-Pandemic label
  annotate(
    "text",
    x = 2020.5,
    y = 20.25,
    label = "POST-PANDEMIC",
    color = "#7a9baa",
    size = 12 / ggplot2::.pt,
    vjust = 0,
    family = "Franklin Gothic"
  ) +

  # Lines
  geom_line(linewidth = 0.9) +

  # Inline labels (no legend)
  geom_text(
    data = tibble(
      year_end = c(2020.5, 2020.5),
      gap_pct  = c(17.6, 13.5),
      label    = c("Mean", "Median"),
      metric   = c("Mean (hourly)", "Median (hourly)")
    ),
    aes(label = label, color = metric),
    hjust = 0,
    size = 4,
    family = "Franklin Gothic",
    show.legend = FALSE
  ) +

  # Scales
  scale_color_manual(values = col_map) +

  scale_x_continuous(
    breaks = df2$year_end,
    expand = expansion(mult = c(0.02, 0))
  ) +

  scale_y_continuous(
    position = "right",
    breaks = seq(10, 22, by = 2),
    expand = expansion(mult = c(0.02, 0.05))
  ) +

  # Custom inside-right y values
  annotate(
    "text",
    x = max(df2$year_end) + 0.25,
    y = y_position,
    label = y_vals,
    hjust = 1,
    size = 12 / ggplot2::.pt,
    color = "gray40",
    family = "Franklin Gothic"
  ) +

  # Titles / caption
  labs(
    title = "Gender Pay Gap at NHS Barts Trust 2018–2021",
    subtitle = "Hourly Pay Gap (%)",
    x = NULL,
    y = NULL,
    caption = "Chart by Annisa Kumaladewi. Source: GOV.UK Gender Pay Gap Service (Barts Health NHS Trust)."
  ) +

  # Theme
  theme(
    aspect.ratio = 3.2 / 7,

    # Default body font
    text = element_text(family = "Miller Text"),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),

    plot.margin = margin(t = 14, r = 24, b = 16, l = 0),

    # Gridlines: horizontal only
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#dcdbd8"),
    panel.grid.minor.y = element_blank(),

    legend.position = "none",

    # Axes
    axis.text.x = element_text(
      family = "Franklin Gothic",
      size = 12,
      color = "gray8"
    ),
    axis.line.x = element_line(color = "gray8"),
    axis.ticks.y = element_blank(),
    axis.text.y  = element_blank(),

    # Title / subtitle
    plot.title = element_text(
      family = "Miller Display",
      face = "bold",
      size = 18,
      hjust = 0,
      margin = margin(b = 25)
    ),
    plot.subtitle = element_text(
      family = "Miller Text",
      size = 16,
      hjust = 0,
      margin = margin(b = 3)
    ),

    # Caption
    plot.caption = element_text(
      family = "Franklin Gothic",
      size = 9,
      color = "#4B4B4B",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

p

```

## Viz 2: Bonus Pay Chart
```{r}
# Turn df into long df
bonus_long <- df2 %>%
  select(
    year_end,
    mean_gap = GenderPayGap_BonusPay_Mean_Percent,
    median_gap = GenderPayGap_BonusPay_Median_Percent
  ) %>%
  pivot_longer(cols = c(mean_gap, median_gap),
               names_to = "metric",
               values_to = "gap_pct") %>%
  mutate(metric = recode(metric,
                         mean_gap = "Mean (bonus)",
                         median_gap = "Median (bonus)"))
```

Labelling
```{r}
# Inline labels
bonus_labels <- tibble(
  year_end = c(2020.5, 2020.5),
  gap_pct  = c(34, 23),                 # <- tweak if needed after you view
  label    = c("Mean", "Median"),
  metric   = c("Mean (bonus)", "Median (bonus)")
)

bonus_col_map <- c("Mean (bonus)" = "#fcaa2d", "Median (bonus)" = "#0ab276")

y_vals_bonus <- seq(15, 45, by = 5)
y_position_bonus <- seq(16, 46, by = 5)
```

```{r}
p_bonus <- ggplot(
  bonus_long,
  aes(x = year_end, y = gap_pct, group = metric, color = metric)
) +

  # Shaded region (post-2020)
  annotate(
    "rect",
    xmin = 2020, xmax = 2021,
    ymin = -Inf, ymax = 45,
    alpha = 0.4, fill = "#e0eaf2"
  ) +

  # Top line + label
  annotate(
    "segment",
    x = 2020, xend = 2021,
    y = 45, yend = 45,
    linewidth = 0.3,
    color = "#7a9baa"
  ) +

  annotate(
    "text",
    x = 2020.5,
    y = 45.4,
    label = "COVID-19",
    color = "#7a9baa",
    size = 12 / ggplot2::.pt,
    vjust = 0,
    family = "Franklin Gothic"
  ) +

  # Y values (inside-right)
  annotate(
    "text",
    x = 2021.23,
    y = y_position_bonus,
    label = y_vals_bonus,
    hjust = 1,
    size = 10 / ggplot2::.pt,
    color = "black",
    family = "Franklin Gothic"
  ) +

  # Lines
  geom_line(linewidth = 0.9) +

  # Inline labels
  geom_text(
    data = bonus_labels,
    aes(label = label),
    hjust = 1,
    size = 4,
    family = "Franklin Gothic",
    show.legend = FALSE
  ) +

  scale_color_manual(values = bonus_col_map) +

  scale_x_continuous(
    breaks = df2$year_end,
    limits = c(2017.8, 2021.28),
    expand = c(0, 0)
  ) +

  scale_y_continuous(
    position = "right",
    breaks = seq(15, 45, by = 5),
    limits = c(15, 46),
    labels = scales::label_number(),
    expand = expansion(mult = c(0.02, 0.04))
  ) +

  labs(
    title = "Gender Bonus Pay Gap",
    subtitle = "Bonus Pay Gap (%)",
    x = NULL, y = NULL
  ) +

  theme(
    # default body
    text = element_text(family = "Miller Text"),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(t = 8, r = 10, b = 10, l = 0),

    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#dcdbd8"),
    panel.grid.minor.y = element_blank(),

    legend.position = "none",

    # X-axis text
    axis.text.x = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 12,
      color = "gray8"
    ),
    axis.line.x = element_line(color = "gray8"),
    axis.ticks.y = element_blank(),
    axis.text.y  = element_blank(),

    # Title / subtitle
    plot.title = element_text(
      family = "Miller Display",
      face = "bold",
      size = 18,
      hjust = 0,
      margin = margin(b = 12)
    ),
    plot.subtitle = element_text(
      family = "Miller Text",
      size = 14,
      hjust = 0,
      margin = margin(b = 3)
    )
  )

p_bonus

```

## Vertical Barbell
```{r}
# Data
bonus_df <- df2 %>%
  select(
    year_end,
    Women = WhoReceivesBonusPay_Women_Percent,
    Men   = WhoReceivesBonusPay_Men_Percent
  ) %>%
  filter(year_end %in% c(2018, 2019, 2020, 2021)) %>%
  mutate(year_end = as.numeric(year_end))

# Long data for points + legend
bonus_pts <- bonus_df %>%
  pivot_longer(cols = c(Women, Men), names_to = "Gender", values_to = "share") %>%
  mutate(Gender = factor(Gender, levels = c("Women", "Men")))

col_gender <- c("Men" = "#4b73d3", "Women" = "#eb535a")

# Y grid / labels inside panel
y_breaks <- 0:4 
y_label_x <- 2021.23 # where the y numbers sit inside the panel (right side)
y_label_position <- seq(0.15, 4.15, by = 1)
```

```{r}
p_barbell_v <- ggplot(bonus_df, aes(x = year_end)) +

  # Post-pandemic shading
  annotate(
    "rect",
    xmin = 2020, xmax = 2021,
    ymin = -Inf, ymax = Inf,
    fill = "#e0eaf2", alpha = 0.4
  ) +

  geom_hline(yintercept = 0, color = "gray8", linewidth = 0.9) +

  # Connector (vertical)
  geom_segment(
    aes(y = Women, yend = Men, xend = year_end),
    color = "#cadfe9",
    linewidth = 2
  ) +

  # Points
  geom_point(
    data = bonus_pts,
    aes(y = share, color = Gender),
    size = 4
  ) +

  # Y labels
  annotate(
    "text",
    x = y_label_x,
    y = y_label_position,
    label = y_breaks,
    hjust = 1,
    vjust = 0.5,
    color = "black",
    size = 12 / ggplot2::.pt,
    family = "Franklin Gothic"
  ) +

  scale_color_manual(
  values = col_gender,
  breaks = c("Women", "Men"),
  labels = toupper(c("Women", "Men"))
) +

  scale_x_continuous(
    breaks = 2018:2021,
    limits = c(2017.8, 2021.28),
    expand = c(0, 0)
  ) +

  scale_y_continuous(
    breaks = y_breaks,
    limits = c(0, 4.3),
    expand = c(0, 0)
  ) +

  labs(
    title = "Percentage of\nWomen vs Men\nReceiving Bonus",
    subtitle = "Share of employees (%)",
    x = NULL, y = NULL
  ) +

  theme(
    # Default text family
    text = element_text(family = "Miller Text"),

    plot.title.position = "plot",
    plot.caption.position = "plot",

    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),

    # horizontal grid only
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#dcdbd8"),
    panel.grid.minor.y = element_blank(),

    # axes
    axis.text.x = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 12,
      color = "gray8"
    ),
    axis.ticks.x = element_line(color = "gray8"),
    axis.ticks.length.x = unit(4, "pt"),
    axis.line.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y  = element_blank(),

    # legend under subtitle
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 12
    ),
    legend.key = element_blank(),
    legend.box.just = "left",
    legend.box.margin = margin(t = -6, b = 8, l = -10),

    # titles
    plot.title = element_text(
      family = "Miller Display",
      face = "bold",
      size = 18,
      hjust = 0,
      margin = margin(b = 10)
    ),
    plot.subtitle = element_text(
      family = "Miller Text",
      size = 14,
      hjust = 0,
      margin = margin(b = 4)
    ),
    plot.caption = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 9,
      color = "gray40",
      hjust = 0,
      margin = margin(t = 10)
    ),

    # margins
    plot.margin = margin(t = 8, r = 10, b = 10, l = 0)
  )

p_barbell_v


```
## Combining chart 2 & 3
```{r}
library(patchwork)

p_top <- p_bonus +
  theme(
    axis.text.x   = element_blank(),
    axis.title.x  = element_blank(),
    plot.caption  = element_blank(),
    plot.margin   = margin(t = 14, r = 10, b = 2, l = 12)
  )

p_bottom <- p_barbell_v +
  theme(
    plot.caption = element_blank(),
    plot.margin  = margin(t = 2, r = 10, b = 10, l = 12)
  )

 p_stacked <-
  (p_top / plot_spacer() / p_bottom) +
  plot_layout(heights = c(1, 0.03, 1))

p_stacked
```
## Stacked Bar Chart Chart: Ethnicity vs Pay Bands
clean
```{r}

wres_df <- read_csv("wres_2020.csv")

wres_median <- wres_df %>%
  summarise(
    s_white   = median(s_white, na.rm = TRUE),
    s_bme     = median(s_bme, na.rm = TRUE),
    s_unknown = median(s_unknown, na.rm = TRUE),

    m_white   = median(m_white, na.rm = TRUE),
    m_bme     = median(m_bme, na.rm = TRUE),
    m_unknown = median(m_unknown, na.rm = TRUE),

    se_white   = median(se_white, na.rm = TRUE),
    se_bme     = median(se_bme, na.rm = TRUE),
    se_unknown = median(se_unknown, na.rm = TRUE),

    vsm_white   = median(vsm_white, na.rm = TRUE),
    vsm_bme     = median(vsm_bme, na.rm = TRUE),
    vsm_unknown = median(vsm_unknown, na.rm = TRUE)
  )

```
wres_long
```{r}
wres_long <- wres_median %>%
  pivot_longer(
    cols = everything(),
    names_to = c("band", "ethnicity"),
    names_sep = "_",
    values_to = "share"
  ) %>%
  mutate(
    band = recode(band,
      s   = "Support",
      m   = "Middle",
      se  = "Senior",
      vsm = "Very senior"
    ),
    ethnicity = recode(ethnicity,
      white   = "White",
      bme     = "BME",
      unknown = "Unknown"
    ),
    band = factor(band, levels = c("Support", "Middle", "Senior", "Very senior"))
  )

```


plot
```{r}
library(scales)

eth_col <- c(
  "White"   = "#70cdde",
  "BME"     = "#fee83e",
  "Unknown" = "#cdcdcd"
)

y_vals  <- seq(0, 100, by = 25)

# Stack order
wres_long$ethnicity <- factor(wres_long$ethnicity, levels = c("White", "BME", "Unknown"))

# Legend order and all caps
legend_breaks <- c("BME", "Unknown", "White")
legend_labels <- toupper(legend_breaks)
```

```{r}
p_wres <- ggplot(wres_long, aes(x = band, y = share, fill = ethnicity)) +
  geom_col(width = 0.6) +

  # Baseline at y = 0 should be black
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6) +

  # Manual fills + legend formatting
  scale_fill_manual(
    values = eth_col,
    breaks = legend_breaks,
    labels = legend_labels
  ) +

  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = c(0, 0),
    limits = c(0, 1.1)
  ) +
  
  # Y-axis labels
  annotate(
    "text",
    x = length(unique(wres_long$band)) + 0.6,  # right side, inside panel
    y = seq(0.05, 1.05, by = 0.25),
    label = y_vals,
    hjust = 1,
    color = "gray10",
    size = 11 / ggplot2::.pt,
    family = "Franklin Gothic"
  ) +


  labs(
    title = "Ethnic composition of NHS staff by seniority",
    subtitle = "Median share across NHS trusts, 2020 (%)",
    x = NULL,
    y = NULL,
    caption = "Chart by Annisa Kumaladewi.\nSource: NHS Workforce Race Equality Standard (WRES)"
  ) +

  theme(
    text = element_text(family = "Miller Text"),

    plot.background  = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    
    plot.margin = margin(t = 8, r = 10, b = 10, l = 0),

    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#d9d9d9", linewidth = 0.6),
    panel.grid.minor.y = element_blank(),

    axis.title = element_blank(),
    axis.ticks = element_blank(),

    axis.text.x = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 12,
      color = "gray10",
      margin = margin(t = 6)
    ),

    axis.text.y = element_blank(),

    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 11,
      color = "gray10"
    ),
    legend.key = element_rect(fill = "white", colour = NA),
    legend.margin = margin(b = 6),
    legend.box.margin = margin(b = 6),

    plot.title.position = "plot",
    plot.title = element_text(
      family = "Miller Display",
      face = "bold",
      size = 20,
      color = "gray10",
      margin = margin(b = 6)
    ),

    plot.subtitle = element_text(
      family = "Miller Text",
      face = "plain",
      size = 14,
      color = "gray20",
      margin = margin(b = 10)
    ),

    plot.caption.position = "plot",
    plot.caption = element_text(
      family = "Franklin Gothic",
      face = "plain",
      size = 9,
      color = "gray40",
      hjust = 0,
      margin = margin(t = 10)
    )
  )

p_wres

```

